---
authors: []
categories: []
date: " "
slides:
  highlight_style: dracula
  theme: black
summary: Week 1, class 1
tags: []
title: Introduction
---


## Goals for semester

- Analyze and interpret general(ized) linear models (aka regression models; aka ANOVA)

- Be able to graph your data

- Identify and diagnose problems when they arise

## Important links
- [Class website](https://josh-jackson.github.io/Psych5067/)

- [Class readings](https://josh-jackson.github.io/Psych5067/readings.html)

- [LSR](https://learningstatisticswithr.com/book/)

- [Example data from class (and Rmd and all other materials)](https://github.com/josh-jackson/Psych5067) if you dont want to use class webpage. 

## Important information

- Most important information is on the slides

- Readings serve to supplement and enhance

- Important equations will be presented via slides

## Two types of people

- This class is going too slow vs going too fast

- If too fast: ASK QUESTIONS AND COME AND VISIT ME. Also spend lots of time outside of class; learning takes time.

- If too slow: ASK QUESTIONS AND COME AND VISIT ME. There are lots of interesting extensions to these models. I have readings. 


## Goals for today  

- Review of last semester
- Overview of this semester

## What should you already know? 


## pragmatics

- R, RStudio 
- Rmarkdown/LaTeX
- Reproducible reports: Rmd/LaTeX -> html/pdf 

- Keep organized: R Projects 
- Keep it visual: ggplot 
- Keeping track/sharing: osf, github (recommended)
- Keep it simple and readable: dplyr & tidyverse (strongly recommended)
- Keep it interesting: shiny apps (recommended for advanced users)

## Visualizing data

- Histograms, probability mass function

- Density plot, probability density function, cumulative density function

- Representing your results through bar, box, violin
1. Avoid wow graphs, 2. use interpretable scales, 3. label 4. always provide a dispersion estimate and 5. if possible include raw data and/or the distribution (as dispersion only can hide oddities)

##

```{r}

# Package for including stats within ggplot and for plotting basic ANOVAs
library(ggstatsplot)

#load data
data("iris")

# plot
ggstatsplot::ggbetweenstats(
  data = iris,
  x = Species,
  y = Sepal.Length,
  xlab = "Type of Species", # label for the x-axis variable
  ylab = "Attribute", # label for the y-axis variable
  title = "Iris flower data set", 
  bf.message = FALSE
) + 
  ggplot2::coord_cartesian(ylim = c(3, 8)) +
  ggplot2::scale_y_continuous(breaks = seq(3, 8, by = 1))
```

## ANOVA results

```{r}
summary(aov(Sepal.Length ~ Species, data = iris))

```

## Welches ANOVA

```{r}

oneway.test(Sepal.Length ~ Species, data = iris, var.equal=FALSE)

```

## Running models

Y = X + E
Dependent variable = Independent Variable(s) + Error
Sum of Squares total = Sums of squares between + sums of squares within

Every model we run this semester will be in the form of Y = X + E. Last semester, every model you ran was in the form of Y = X + E. You just didn't know it. 

##

```{r, eval=FALSE}

t.1 <- t.test(y ~ x, data = d) 
# y is cont and x is a categoriocal/nominal (dichotomous) factor
```

## Comparing models

SSW(restricted) - SSW(full) / stuff

The full restricted is equivalent to Y = X + E compared to Y = E and asking which E is larger. The only difference between these two models is our X variable (our IVs). The result is we are asking whether X helps explain variance in Y, which is equivalent to asking whether X reduces variance that is unexplained. 

## Theoretical probability distributions 

- How you measure your variables determine your analytic tool options

- For IVs how you measure changes the name of the analysis but not really much else

- For DVs, data generating process (DGP) defines appropriate analyses 

- Binomial, poisson, negative binomial, normal (Gaussian), t, F, chi-square...

- As a default we assume Gaussian. Why? 

- THIS DOES NOT MEAN WE NEED TO HAVE A NORMAL DISTRIBUTION IN OUR DATA

## Many distributions

- Population distribution -> sample distribution -> sampling distrinbution

- Standard error is the standard deviation of the sampling distribution

## p-values, NHST and standard opperating procedures 

- p(D|Ho), not p(Ho|D) 

- Type 1 & 2 errors, power, family wise error

- Issues with NHST

- Questionable research practices (QRPs)

## Effect size and CIs
- There is imprecision in all that you do. Document it. 

- Interpretation and utility of CI?

- Common effect size metrics?

- How does the CI relate to the sample and sampling distribution? 

- Engage in meta-analytic thinking

## how to suceed (in class)

- Go through slides, work through code
- Apply concepts to your own work
- Devote time

## Looking forward

-  This semeter's themes are Relationships and Models





