---
title: "Introduction"
author: Josh Jackson
date: "1-14-20"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    seal: false
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: "16:9"
---

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
mono_light(
  base_color = "#23395b",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Droid Mono")
)
```

<style type="text/css">
.remark-slide-content {
    font-size: 30px;
    padding: 1em 4em 1em 4em;
}
</style>


# Goals for semester

- Analyze and interpret general(ized) linear models (aka regression models; aka ANOVA)

- Be able to graph your data

- Identify and diagnose problems when they arise

---
# Important links

- [LSR](https://learningstatisticswithr.com/book/)

- [Example data from class (and Rmd and all other materials)](https://github.com/josh-jackson/5067-Spring-2020) if you don't want to use class webpage. 

---
# Important information

- Most important information is on the slides

- Readings serve to supplement and enhance

- Important equations will be presented via slides

---
# Two types of people

- This class is going too slow vs going too fast

- If too fast: ASK QUESTIONS AND COME AND VISIT ME. Also spend lots of time outside of class; learning takes time.

- If too slow: ASK QUESTIONS AND COME AND VISIT ME. There are lots of interesting extensions to these models. I have readings. 

---
# Goals for today  

- Review of last semester

- Overview of this semester

---
# What should you already know? 

---
# Pragmatics

- R, RStudio 
- Rmarkdown/LaTeX
- Reproducible reports: Rmd/LaTeX -> html/pdf 
- Keep organized: R Projects
- Keep it visual: ggplot
- Keeping track/sharing: osf, github (recommended)
- Keep it simple and readable: dplyr & tidyverse (strongly recommended)
- Keep it interesting: shiny apps (recommended for advanced users)

---
# Visualizing data

- Histograms, probability mass function

- Density plot, probability density function, cumulative density function

- Representing your results through bar, box, violin

- Suggestions: 1. Avoid wow graphs, 2. use interpretable scales, 3. label 4. always provide a dispersion estimate and 5. if possible include raw data and/or the distribution (as dispersion only can hide oddities)

---
class: clear
.code60[
```{r, message=FALSE}
# Package for including stats within ggplot and for plotting basic ANOVAs
library(ggstatsplot)

#load data
data("iris")

# plot
P1.1 <- ggstatsplot::ggbetweenstats(
data = iris,
  x = Species,
  y = Sepal.Length,
  xlab = "Type of Species", # label for the x-axis variable
  ylab = "Attribute", # label for the y-axis variable
  title = "Iris flower data set", 
  bf.message = FALSE
) + 
  ggplot2::coord_cartesian(ylim = c(3, 8)) +
  ggplot2::scale_y_continuous(breaks = seq(3, 8, by = 1))
```
]

---
class: clear
```{r dpi = 300, out.height=600, out.width = 1000,echo=FALSE}
P1.1
```


---
class: clear
ANOVA vs Welch's ANOVA
.code70[
```{r}
summary(aov(Sepal.Length ~ Species, data = iris))

```


```{r}

oneway.test(Sepal.Length ~ Species, data = iris, var.equal=FALSE)

```
]

---
# Running models

Y = X + E 

Dependent variable = Independent Variable(s) + Error 

Sum of Squares total = Sums of squares between + sums of squares within 

Every model we run this semester will be in the form of Y = X + E.  Last semester, every model you ran was in the form of Y = X + E. You just didn't know it. 

```{r, eval=FALSE}

t.1 <- t.test(y ~ x, data = d) 
# y is cont and x is a categoriocal/nominal (dichotomous) factor
```

---
# Comparing models

SSW(restricted) - SSW(full) / stuff 

The full restricted is equivalent to Y = X + E compared to Y = E and asking which E is larger. The only difference between these two models is our X variable (our IVs). The result is we are asking whether X helps explain variance in Y, which is equivalent to asking whether X reduces variance that is unexplained. 
---
# Theoretical probability distributions 

- How you measure your variables determine your analytic tool options

- For IVs how you measure changes the name of the analysis but not really much else

- For DVs, data generating process (DGP) defines appropriate analyses 

- Binomial, poisson, negative binomial, normal (Gaussian), t, F, chi-square...

- As a default we assume Gaussian. Why? 

- THIS DOES NOT MEAN WE NEED TO HAVE A NORMAL DISTRIBUTION IN OUR DATA
---
# Many distributions

- Population distribution -> sample distribution -> sampling distribution

- Standard error is the standard deviation of the sampling distribution
---

# p-values, NHST and standard operating procedures 

- p(D|Ho), not p(Ho|D) 

- Type 1 & 2 errors, power, family wise error

- Issues with NHST

- Questionable research practices (QRPs)

---
# Effect size and CIs
- There is imprecision in all that you do. Document it. 

- Interpretation and utility of CI?

- Common effect size metrics?

- How does the CI relate to the sample and sampling distribution? 

- Engage in meta-analytic thinking

---
# How to succeed (in class)

- Go through slides, work through code  

- Apply concepts to your own work 

- Devote time 

---
# Looking forward

-  This semester's themes are Relationships and Models





